\Large Last Writer Slicing: Efficient Data Provenance for Concurrent Program Debugging and Analysis

### 2013-11-08 paper.tex
committing some stuff before a break

Debugging a concurrent program requires understanding how threads interact and
{\em data provenance information} furnishes that understanding. A value's
provenance is a description of the reason it is held by a variable.  Provenance
can take on many forms, for example, a value's provenance could be the
instruction that wrote it or the function that returned it.  Provenance
information helps answer the question: "why does this variable contain this
value?".  During debugging a programmer can use that information to connect 
disparate parts of a program that interact, leading to a failure.

@@

Prior work has shown that provenence helps with debugging~\cite{badapples,
whylineicse, tipslicingsurvey}.  One effort tracked code that wrote unusable
values into variables~\cite{badapples}.  That work showed that revealing the
connection between a bad value's origin and a failure it causes is helpful
during debugging.  Other work showed that programmers debug programs more
effectively when they can explicitly ask "why" questions during
debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown the promise of provenance, but are not a panacea.
First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}.  Limiting the
applicability of their technique to only some failures, like {\tt Null Pointer
Exceptions}.  Second, while prior work has shown that limited provenance
tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight memory
checkers~\cite{badapples} impose overheads too high for production for C and
C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Focusing on
concurrency is important because a value could have provenance in any of a
large number of threads, making reasoning about why a variable has a value
complex and difficult.

@@

Last writer slices are primarily useful for concurrency debugging, but we also
observe that provenance information in the form of last writer slices is a
useful foundation for implementing other concurrent program analyses.  Several
tools from prior work has used {\em ad hoc} last-writer information as part of
more heavy-weight, sophisticated in domains like bug detection ({\em e.g.},
~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
~\cite{recon,conmem,conseq,cci}), profiling ({\em
e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
understanding~\cite{oshatr}.

Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
Traps} or {\em \ctraps}  on top of last writer slices
(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each potentially
shared memory location when threads access those locations.  If the accessing thread
is different from the thread that last wrote, the access signfies {\em communication}
between the threads.  Communication is a common subject of

monitors inter-thread communication with
overhead low enough for production use in many important cases.  When threads
communicate, \ctraps delivers a software {\em trap} that can be handled by a
trap handler.  Trap handlers implement {\em \ctraps applications} through an
extensible interface that can analyze or interpose on the execution
(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
applications can perform arbitrary dependence analyses, but at their own
discretion and cost.  As we show in our evaluation, the overhead of \ctraps
applications scales with the complexity of their analysis.

@@

\item{We describe last writer slicing, a dynamic analysis that tracks
memory updates, capturing data provenance information.  We
design and implement system support for collecting last writer slices.  }

\item{Building on last writer slicing, we describe and implement \ctraps, which
exposes inter-thread communication through an extensible trap interface.  }

\item{We evaluate the utility of last writer slices and \ctraps. We first
present several case studies that show that last writer slices are useful for
debugging real-world bugs.  We then show how \ctraps is useful for building
dependence analyses by implementing two analyses from prior
work~\cite{cci,defuse,recon}. }

>>>>

Debugging a concurrent program requires understanding how threads interact and
{\em data provenance information} furnishes that understanding. A value's
provenance is a description of the reason that it is stored in a variable.
Provenance can take on many forms, for example, a value's provenance could be
the instruction that wrote it or the function that returned it.  Provenance
information helps answer the question: "Why does this variable contain this
value?", which is central to debugging.

%During debugging provenance helps a programmer better understand not just where failures occur, but where values came from, connectng disparate
%parts of a program that interact, leading to a failure.

@@

Prior work has shown that provenence helps with debugging~\cite{badapples,
whylineicse, tipslicingsurvey}.  One effort tracked code that wrote unusable
values into variables~\cite{badapples}.  That work showed that revealing the
connection between a bad value's origin and a failure it causes is helpful
during debugging.  Other work showed that programmers debug programs more
effectively when they can explicitly ask "why" questions about the origin of
certain values during debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea.  First, prior work~\cite{badapples} has focused on providing
provenance information for only some unusable {\em values}, like {\tt NULL}.
Limiting the applicability of their technique to only some failures, like {\tt
Null Pointer Exceptions}.  Second, while prior work has shown that limited
provenance tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples} impose overheads too high for production for C
and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Focusing on
concurrency is important because a value could have provenance in any of a
large number of threads, making reasoning about why a variable has a value
complex and difficult.

@@

Last writer slices are primarily useful for concurrency debugging, but we also
observe that provenance information in the form of last writer slices is a
useful foundation for implementing other concurrent program analyses.  Several
tools from prior work has used {\em ad hoc} last-writer information as part of
more heavy-weight analyses in domains like bug detection ({\em e.g.},
~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
~\cite{recon,conmem,conseq,cci}), profiling ({\em
e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
understanding~\cite{oshatr}.

Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
Traps} or {\em \ctraps}  on top of last writer slices
(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each
potentially shared memory location when a thread accesses such a location.  If
the accessing thread is different from the thread that last wrote the location,
the access signifies {\em communication} between the threads.  When
communication occurs, \ctraps delivers a software {\em trap} to the accessing
thread giving it information about how communication occurred.  The trap can be
handled by an analysis-specific trap handler.  Trap handlers implement {\em
\ctraps applications} through an extensible interface that can interpose on the
execution (Section~\ref{sec:apps}).  By default, \ctraps has very low overheads
and similarly to last writer slicing are appropriate for use in production in
many cases.  \ctraps applications can perform arbitrary concurrency analyses,
but at their own discretion and cost.  As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.

@@

\item{We describe last writer slicing, an novel data provenance tracking
mechanism, that produces information that is useful for debugging concurrent
programs.  We design and implement system support for collecting last writer
slices that is efficient enough for use in production.  }

\item{Building on last writer slicing, we develop \ctraps, an extensible
framework for implementing analyses of inter-thread communication in
multi-threaded programs.  }

\item{We evaluate the utility of last writer slices and \ctraps. We first
present a study of several real concurrent program bugs and show that last
writer slices are useful for debugging.  We then show how \ctraps is useful for
building dependence analyses by implementing two analyses from prior
work~\cite{cci,defuse,recon}. }
